Free Theorems for Nested Types

### 2020-07-01 free-theorems-nesteds/paper5.tex
small changes/typos to intro/abstract

\begin{abstract}
This paper considers parametricity and its consequent free theorems
for nested data types. Rather than representing nested types via their
Church encodings into a higher-kinded extension of System F, we adopt
a functional programming perspective and design a polymorphic calculus
that provides primitives for constructing a robust class of nested
types directly as fixpoints.
%\`{a} la Pitts' PolyPCF.
At the term level, our calculus supports
%the usual
pattern matching, map functions, and stylized recursion via folds for
all such types.
%Using a variation on Atkey's 2012 construction for $F_\omega$,
%
%of a parametric model
%that explicitly constructs equality operators rather than merely
%postulating them,
Our main technical contribution is the construction of a parametric
model for this calculus. This is both delicate and challenging.  To
establish a suitable Identity Extension Lemma for our calculus, extra
care must be taken in the formation of fixpoint types; all ADTs and
all (truly) nested types appearing in the literature can nevertheless
be expressed in our calculus. In addition, to ensure the existence of
semantic fixpoints interpreting nested types, cocontinuity conditions
on the functors interpreting all of the types in our calculus must
be threaded throughout our model construction.  Overall, this paper
provides a blueprint for deriving principled parametricity results for
practical programming languages, such as Haskell and Agda, that
directly construct nested types as fixpoints rather than representing
them via Church encodings.
%encoding them.
It also shows how parametric models can derive useful free
theorems for programs over nested types. Like their counterparts for
ADTs, these free theorems
%our model derives
provide
%, in particular,
a principled alternative to proving program properties via induction,
e.g., using the deep induction rules for nested types recently
developed by Johann and Polonsky.
%We derive a number of free theorems for nested types.
%To illustrate, we derive a number of free theorems.
Our approach is illustrated with several examples. 
\end{abstract}

@@

{\small
\[\begin{array}{l}
\mathtt{data\; List \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{nil\;:\; List\;A}\\
\hspace*{0.4in}\mathtt{cons\;:\;A \rightarrow List\;A \rightarrow List\;A}
\end{array}\]}
A natural approach to the problem uses structural induction on the
input data structure in question. This requires knowing not just the
definition of the ADT of which the input data structure is an
instance, but also the program text for the functions involved in the
properties to be proved. For example, to prove by induction that
mapping a polymorphic function over a list and then reversing the
resulting list is the same as reversing the original list and then
mapping the function over the result, we unwind the (recursive)
definitions of the $\mathtt{reverse}$ and $\mathtt{map}$ functions
over lists to according to the inductive structure of the input
list. Such data-driven induction proofs over ADTs are routine enough
to be included in, say, an undergraduate functional programming
course.

@@

Unfortunately, it is nowhere near as clear that analogous intuitive or
formal inductive arguments can be made for the
$\mathtt{map}$-$\mathtt{reverse}$ property for bushes. Indeed, a proof
by induction on the input bush must recursively induct over the bushes
that are internal to the top-level bush in the recursive position.
This is sufficiently delicate that no induction rule for bushes or
other truly nested types was known until very recently, when {\em deep
  induction}~\cite{jp20} was developed as a way to induct over {\em
  all} of the structured data present in an input. Deep induction thus
not only gave the first principled and practically useful structural
induction rules for bushes and other truly nested types, and has also
opened the way for incorporating automatic generation of such rules
for (truly) nested data types --- and, eventually, even GADTs --- into
modern proof assistants.
\begin{figure*}
\hspace*{-0.7in}
\resizebox{0.35\linewidth}{!}{
\begin{minipage}[t]{0.5\textwidth}
\[\begin{array}{l}
\mathtt{reversePTree : \forall \{A : Set\}  \rightarrow PTree\; A
  \rightarrow PTree \;A}\\  
\mathtt{reversePTree \; \{A\} = foldPTree \; \{A\}\; \{PTree\}}\\
\hspace*{1.3in} \mathtt{pleaf}\\
\hspace*{1.3in} \mathtt{( \lambda p \rightarrow pnode\; (mapPTree\;
  swap\; p) )}\\  
\\
\mathtt{foldPTree : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow Set \}
  \rightarrow}\\
\hspace*{0.8in} \mathtt{( \{B : Set\} \rightarrow B \rightarrow F B) \rightarrow}\\
\hspace*{0.8in} \mathtt{(\{B : Set\} \rightarrow F (B \times B)
  \rightarrow F B) \rightarrow}\\
\hspace*{0.8in}\mathtt{PTree \;A \rightarrow F\;A}\\
\mathtt{foldPTree\; n\; c\; (pleaf\; x) = n \;x}\\
\mathtt{foldPTree\; n\; c \;(pnode\; p) = c\; (foldPTree \;n \;c \;p)}\\
\\
\mathtt{mapPTree : \forall \{A\, B : Set\} \rightarrow (A \rightarrow B)
  \rightarrow PTree\; A \rightarrow PTree\; B}\\ 
\mathtt{mapPTree \;f\; (pleaf \;x) = pleaf\; (f\; x)}\\
\mathtt{mapPTree\; f \;(pnode\; p) =}\\
\hspace*{0.7in}  \mathtt{pnode\; (mapPTree \;(\lambda p
  \rightarrow (f (\pi_1 \,p), f (\pi_2\,p)))\; p)}\\ 
\\
\mathtt{swap : \forall \{A : Set\} \rightarrow (A \times A)
  \rightarrow (A \times A)}\\ 
\mathtt{swap \;(x,y) = (y,x)}
\end{array}\]
\caption{$\mathtt{reversePTree}$ and auxiliary functions in Agda}\label{fig:funs} 
\end{minipage}}
\quad\quad\quad\quad\quad%\quad\quad\quad
\resizebox{0.35\linewidth}{!}{
\begin{minipage}[t]{0.5\textwidth}
\[\begin{array}{l}
  \mathtt{reverseBush : \forall \{A : Set\} \rightarrow Bush\; A
  \rightarrow Bush\; A}\\ 
\mathtt{reverseBush \;\{A\} = foldBush\; \{A\}\; \{Bush\}\; bnil \;balg}\\
\\
\mathtt{foldBush : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow
  Set\} \rightarrow}\\
\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow F B) \rightarrow}\\
\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow B \rightarrow F\;
  (F\; B) \rightarrow F\; B) \rightarrow}\\ 
\hspace*{0.5in} \mathtt{Bush\; A \rightarrow F \;A}\\
\mathtt{foldBush\; bn\; bc\; bnil = bn}\\
\mathtt{foldBush \;bn \;bc \;(bcons\; x \;bbx) =}\\
\hspace*{0.5in} \mathtt{bc\; x \;(foldBush \;bn\; bc
  \;(mapBush \;(foldBush\; bn\; bc) \;bbx))}\\
\\
\mathtt{mapBush : \forall \{A \,B : Set\} \to (A \to B) \to Bush\, A \to
  Bush\, B}\\ 
\mathtt{mapBush\; \_ \; bnil = bnil}\\
\mathtt{mapBush\; f\; (bcons\; x \;bbx) = bcons \;(f\, x) \; (mapBush\;
  (mapBush\, f) \; bbx)}\\ 
\\
\mathtt{balg : \forall \{B : Set\} \rightarrow B \rightarrow Bush
  \;(Bush\; B) \rightarrow Bush\; B}\\
\mathtt{balg \;x \;bnil = bcons\; x \;bnil}\\
\mathtt{balg \;x\; (bcons \;bnil \;bbbx) = bcons\; x \;(bcons \;bnil\; bbbx)}\\
\mathtt{balg\; x \;(bcons \;(bcons\; y \;bbx)\; bbbx) =}\\
\hspace*{0.5in} \mathtt{bcons\; y\; (bcons\; (bcons \;x \;bbx) \;bbbx)}\\
\end{array}\]
\caption{$\mathtt{reverseBush}$ and auxiliary
  functions in Agda}\label{fig:funs2} 
\end{minipage}}
\end{figure*}

@@

variable contexts to set or relations, as appropriate. But to ensure

@@

we derive several of this latter kind of free theorems from it in

@@

We are not the first to consider parametricity at higher
kinds. Atkey~\cite{atk12} constructs a parametric model for full
System $F_\omega$, but within the impredicative Calculus of Inductive
Constructions (iCIC) rather than in a semantic category. His
construction is in some ways similar to ours, but it represents data
types using Church encodings rather than constructing them via
primitives. Since his model is entirely syntactic, his syntactic
``functors'', whose associated $\mathit{fmap}$ functions representing
their functorial actions must be {\em given} together with their
underlying type constructors, cannot be interpreted as semantic
functors. The associated Church encodings therefore cannot be
interpreted as their fixpoints, so Atkey need not, and does not,
impose cocontinuity conditions on his model to ensure that these
fixpoints exist.  Nevertheless, Atkey does verify the existence of
initial algebras for ``functors'' in iCIC. He does not indicate which
type constructors support the kinds of $\mathit{fmap}$ functions
needed to ensure they are ``functors'', but we suspect spelling this
out explicitly would result in a full higher-kinded extension of the
calculus presented here.

\section{The Calculus}\label{sec:calculus}

>>>>

\begin{abstract}
This paper considers parametricity and its consequent free theorems
for nested data types. Rather than representing nested types via their
Church encodings in a higher-kinded extension of System F, we adopt
a functional programming perspective and design a polymorphic calculus
that provides primitives for constructing a robust class of nested
types directly as fixpoints.
%\`{a} la Pitts' PolyPCF.
At the term level, our calculus supports
%the usual
pattern matching, map functions, and stylized recursion via folds for
all such types.
%Using a variation on Atkey's 2012 construction for $F_\omega$,
%
%of a parametric model
%that explicitly constructs equality operators rather than merely
%postulating them,
Our main technical contribution is the construction of a parametric
model for this calculus. This is both delicate and challenging.  To
establish a suitable Identity Extension Lemma for our calculus, extra
care must be taken in the formation of fixpoint types; all ADTs and
all (truly) nested types appearing in the literature can nevertheless
be expressed in our calculus. In addition, to ensure the existence of
semantic fixpoints interpreting nested types, cocontinuity conditions
on the functors interpreting all of the types in our calculus must
be threaded throughout our model construction.  Overall, this paper
provides a blueprint for deriving principled parametricity results for
practical programming languages, such as Haskell and Agda, that
directly construct nested types as fixpoints rather than representing
them via Church encodings.
%encoding them.
It also shows how parametric models can derive useful free
theorems for programs over nested types. Like their counterparts for
ADTs, these free theorems
%our model derives
provide
%, in particular,
a principled alternative to proving program properties via induction,
e.g., using the deep induction rules for nested types recently
developed by Johann and Polonsky.
%We derive a number of free theorems for nested types.
%To illustrate, we derive a number of free theorems.
Our approach is illustrated with several examples. 
\end{abstract}

@@

{\small
\[\begin{array}{l}
\mathtt{data\; List \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{nil\;:\; List\;A}\\
\hspace*{0.4in}\mathtt{cons\;:\;A \rightarrow List\;A \rightarrow List\;A}
\end{array}\]}
A natural approach to the problem uses structural induction on the
input data structure in question. This requires knowing not just the
definition of the ADT of which the input data structure is an
instance, but also the program text for the functions involved in the
properties to be proved. For example, to prove by induction that
mapping a polymorphic function over a list and then reversing the
resulting list is the same as reversing the original list and then
mapping the function over the result, we unwind the (recursive)
definitions of the $\mathtt{reverse}$ and $\mathtt{map}$ functions
over lists according to the inductive structure of the input
list. Such data-driven induction proofs over ADTs are routine enough
to be included in, say, an undergraduate functional programming
course.

@@

Unfortunately, it is nowhere near as clear that analogous intuitive or
formal inductive arguments can be made for the
$\mathtt{map}$-$\mathtt{reverse}$ property for bushes. Indeed, a proof
by induction on the input bush must recursively induct over the bushes
that are internal to the top-level bush in the recursive position.
This is sufficiently delicate that no induction rule for bushes or
other truly nested types was known until very recently, when {\em deep
  induction}~\cite{jp20} was developed as a way to induct over {\em
  all} of the structured data present in an input. Deep induction thus
not only gave the first principled and practically useful structural
induction rules for bushes and other truly nested types, but has also
opened the way for incorporating automatic generation of such rules
for (truly) nested data types --- and, eventually, even GADTs --- into
modern proof assistants.
\begin{figure*}
\hspace*{-0.7in}
\resizebox{0.35\linewidth}{!}{
\begin{minipage}[t]{0.5\textwidth}
\[\begin{array}{l}
\mathtt{reversePTree : \forall \{A : Set\}  \rightarrow PTree\; A
  \rightarrow PTree \;A}\\  
\mathtt{reversePTree \; \{A\} = foldPTree \; \{A\}\; \{PTree\}}\\
\hspace*{1.3in} \mathtt{pleaf}\\
\hspace*{1.3in} \mathtt{( \lambda p \rightarrow pnode\; (mapPTree\;
  swap\; p) )}\\  
\\
\mathtt{foldPTree : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow Set \}
  \rightarrow}\\
\hspace*{0.8in} \mathtt{( \{B : Set\} \rightarrow B \rightarrow F\;B) \rightarrow}\\
\hspace*{0.8in} \mathtt{(\{B : Set\} \rightarrow F\;(B \times B)
  \rightarrow F \;B) \rightarrow}\\
\hspace*{0.8in}\mathtt{PTree \;A \rightarrow F\;A}\\
\mathtt{foldPTree\; n\; c\; (pleaf\; x) = n \;x}\\
\mathtt{foldPTree\; n\; c \;(pnode\; p) = c\; (foldPTree \;n \;c \;p)}\\
\\
\mathtt{mapPTree : \forall \{A\, B : Set\} \rightarrow (A \rightarrow B)
  \rightarrow PTree\; A \rightarrow PTree\; B}\\ 
\mathtt{mapPTree \;f\; (pleaf \;x) = pleaf\; (f\; x)}\\
\mathtt{mapPTree\; f \;(pnode\; p) =}\\
\hspace*{0.7in}  \mathtt{pnode\; (mapPTree \;(\lambda p
  \rightarrow (f (\pi_1 \,p), f (\pi_2\,p)))\; p)}\\ 
\\
\mathtt{swap : \forall \{A : Set\} \rightarrow (A \times A)
  \rightarrow (A \times A)}\\ 
\mathtt{swap \;(x,y) = (y,x)}
\end{array}\]
\caption{$\mathtt{reversePTree}$ and auxiliary functions in Agda}\label{fig:funs} 
\end{minipage}}
\quad\quad\quad\quad\quad%\quad\quad\quad
\resizebox{0.35\linewidth}{!}{
\begin{minipage}[t]{0.5\textwidth}
\[\begin{array}{l}
  \mathtt{reverseBush : \forall \{A : Set\} \rightarrow Bush\; A
  \rightarrow Bush\; A}\\ 
\mathtt{reverseBush \;\{A\} = foldBush\; \{A\}\; \{Bush\}\; bnil \;balg}\\
\\
\mathtt{foldBush : \forall \{A : Set\} \rightarrow \{F : Set \rightarrow
  Set\} \rightarrow}\\
\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow F \;B) \rightarrow}\\
\hspace*{0.5in} \mathtt{(\{B : Set\} \rightarrow B \rightarrow F\;
  (F\; B) \rightarrow F\; B) \rightarrow}\\ 
\hspace*{0.5in} \mathtt{Bush\; A \rightarrow F \;A}\\
\mathtt{foldBush\; bn\; bc\; bnil = bn}\\
\mathtt{foldBush \;bn \;bc \;(bcons\; x \;bbx) =}\\
\hspace*{0.5in} \mathtt{bc\; x \;(foldBush \;bn\; bc
  \;(mapBush \;(foldBush\; bn\; bc) \;bbx))}\\
\\
\mathtt{mapBush : \forall \{A \,B : Set\} \to (A \to B) \to Bush\, A \to
  Bush\, B}\\ 
\mathtt{mapBush\; \_ \; bnil = bnil}\\
\mathtt{mapBush\; f\; (bcons\; x \;bbx) = bcons \;(f\, x) \; (mapBush\;
  (mapBush\, f) \; bbx)}\\ 
\\
\mathtt{balg : \forall \{B : Set\} \rightarrow B \rightarrow Bush
  \;(Bush\; B) \rightarrow Bush\; B}\\
\mathtt{balg \;x \;bnil = bcons\; x \;bnil}\\
\mathtt{balg \;x\; (bcons \;bnil \;bbbx) = bcons\; x \;(bcons \;bnil\; bbbx)}\\
\mathtt{balg\; x \;(bcons \;(bcons\; y \;bbx)\; bbbx) =}\\
\hspace*{0.5in} \mathtt{bcons\; y\; (bcons\; (bcons \;x \;bbx) \;bbbx)}\\
\end{array}\]
\caption{$\mathtt{reverseBush}$ and auxiliary
  functions in Agda}\label{fig:funs2} 
\end{minipage}}
\end{figure*}

@@

variable contexts to sets or relations, as appropriate. But to ensure

@@

we derive several of this latter kind of free theorem from it in

@@

We are not the first to consider parametricity at higher
kinds. Atkey~\cite{atk12} constructs a parametric model for full
System $F_\omega$, but within the impredicative Calculus of Inductive
Constructions (iCIC) rather than in a semantic category. His
construction is in some ways similar to ours, but it represents data
types using Church encodings rather than constructing them via
primitives. However, the $\mathit{fmap}$ functions associated to 
Atkey's functors must be {\em given} together with their
underlying type constructors, and he need not 
impose cocontinuity conditions on his model to ensure that 
fixpoints of his functors exist. Unfortunately, Atkey does not 
indicate which
type constructors support the $\mathit{fmap}$ functions
needed for them to be functors, but we suspect spelling this
out explicitly would result in a full higher-kinded extension of the
calculus presented here.
\section{The Calculus}\label{sec:calculus}
